Hyperplane Elimination for Quickly Enumerating Local Optima

### 2015-09-24 Hyperplane Enumeration/main.tex
Now under the page limit

\begin{abstract}
Examining the properties of local optima is a common method for understanding
combinatorial-problem landscapes.
Unfortunately,
exhaustive methods for finding local optima are limited to very small problem sizes.
We propose a method for exploiting problem structure
to skip hyperplanes
that cannot contain local optima. In most of our experiments,
this hyperplane elimination allows enumeration time to scale with
the total number of local optima instead of with the total landscape size.
We prove optimality for linear and $k$-bounded problems, and we provide
empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
Furthermore, we are able to refine this method to find only solutions
that cannot be improved by flipping $r$ or fewer bits. Counterintuitively,
increasing $r$ can reduce total runtime by drastically reducing the
number of local optima. While previous methods were limited to
landscapes with at most $2^{30}$ binary strings, hyperplane elimination can enumerate the same problems with
$2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
We believe this increase in size will help further our understanding of complex real-world landscapes.

@@

We use two common variants of NKq in our experiments which specify how the $K$ additional
variables in each subfunction are chosen: Adjacent and Random. In Adjacent NKq, $f_i$ depends
on variable indices $[i..(i+k) \bmod N]$. In Random NKq, $f_i$ depends on $x_i$ and a random
set of $K$ unique variables that does not include $x_i$. While Random NKq is NP-Hard,
the structure of Adjacent NKq allows for a polynomial time dynamic programming solution~\cite{wright:2000:solvingnk}.
Adjacent NKq is therefore often used to understand a search algorithm's ability to reach the known global
optimum.

@@

The first major result in Gray-Box optimization was the proof that the list
of fitness-improving moves from a solution can be updated after a bit flip in \BigO{1}
time~\cite{whitley:2012:constant}.
The fitness effect of flipping $x_j$ changes after flipping $x_i$ only if
there is a non-linear relationship between $x_i$ and $x_j$.
In an Mk Landscape, variables
can only have a non-linear relationship if they appear together in at least one $s_i$.
Therefore by definition the total number of non-linear
relationships in an Mk Landscape is linear with $N$, meaning on average the amortized
number of non-linear relationships per variable is \BigO{1}.

This result was extended by~\cite{chicano:2014:ball} to include local search for local
optima that cannot be improved using $r$ or fewer bits with no increase in asymptotic costs.
Consider two variables $x_i$ and $x_j$ that do not appear in the same $s_i$.
By definition the fitness effect of flipping both $x_i$ and $x_j$ is equal to the sum
of flipping each independently. As a result, if neither individual flip is fitness improving,
flipping both together cannot be fitness improving. By examining the graph of non-linear
interactions between variables, this principle can be extended to any collection of $r$
variables, with the number of useful collections growing at \BigO{N} when $r$ is a small constant.
As a result, the time to update the list of improving moves after up to $r$ bit-flips is still \BigO{1}.

@@

\section{Hyperplane Elimination}
\begin{figure}
  \centering
  \includegraphicsfit{Enumerate}
  \caption{Example change of ordering. The gray variables represent all dependencies
           for move $m_i$ which flips variable $F$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
  \label{fig-enumerate}
\end{figure}

Due to the limited non-linearity of the Gray-Box domain, it is possible to exclude large
parts of the search space without missing any local optima.
Consider the representation presented in the top of Figure~\ref{fig-enumerate}. In a Black-Box
domain, enumeration would progress as a binary counter, treating $index$ zero (symbol $A$ in the solution) as
the least significant bit. This ordering ensures that before changing $index$ $i$, all possible settings of $index$
0 through $i-1$ have been tested. This corresponds to examining the hyperplane where the lowest $i-1$ positions
can vary and all other positions remain fixed. The Gray-Box domain makes it possible to skip hyperplanes
that cannot contain local optima. In Figure~\ref{fig-enumerate}, move $m_i$ which flips variable $F$ is a fitness improvement
when enumeration starts (all variables set to 0). Due to the known relationships between variables,
we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
Therefore, until one of those four variables are modified, the solution cannot be a local optimum.
As a result, the hyperplane **0*00*0 cannot contain a local optima and can
therefore be eliminated from consideration during enumeration.
More generally, if at any point during enumeration
there exists a fitness-improving move, no local optima can exist until at least one
dependency of that move is modified. Any hyperplane that has all of a fitness-improving
move's dependencies fixed cannot contain any local optima.

\begin{algorithm}
  \caption{Find all local optima using Hyperplane Elimination.}
  \label{alg-enumerate}
  \begin{algorithmic}[1]
    \Require $solution \leftarrow \{0\}^N$
    \Require $move\_bin$ (List of sets of improving moves in $solution$)
    \Require \textproc{MakeMove} (Function that flips $index$ in $solution$ and updates $move\_bin$)
    \Ensure $found$ (list of all local optima)
    \State $found \leftarrow [~]$
    \State $index \leftarrow N-1$
    \While{$index < N$}
      \While{$index > 0$ \textbf{and} $move\_bin[index] = \emptyset$}\label{alg-enumerate-bincheck}
        \State $index \leftarrow index-1$
      \EndWhile
      \If{$index = -1$}
        \State $found \leftarrow found + [solution]$
        \State $index \leftarrow 0$
      \EndIf
      \While{$index < N$ \textbf{and} $solution[index] = 1$}\label{alg-enumerate-counter}
        \State \Call{MakeMove}{$index$}
        \State $index \leftarrow index + 1$
      \EndWhile
      \If{$index < N$}
        \State \Call{MakeMove}{$index$}\label{alg-enumerate-insert}
      \EndIf
    \EndWhile
  \end{algorithmic}
\end{algorithm}

@@

\section{Reordering Variables}
\label{sec-reorder}
As a final efficiency, the order in which variables in $solution$ are $index$ed can be reordered.
When a move is fitness-improving,
the amount of search space that is skipped depends on how high its lowest $index$ dependency is. Therefore,
by rearranging the order to make its lowest $index$ dependency higher, more search space can be skipped. 
Figure~\ref{fig-enumerate} shows how changing the $index$ order of variables
improves $m_i$'s lowest $index$ dependency from 2 to 4. Consider that before reordering $m_i$ allows the hyperplane
**000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering the hyperplane that can be skipped is ****0000.
Now, whenever $m_i$ is a fitness improvement, Line~\ref{alg-enumerate-bincheck} in Algorithm~\ref{alg-enumerate}
skips 4 times as many solutions.

We perform this reordering in a greedy fashion, such that the move with the
least-unmapped dependencies has all of its remaining dependencies mapped to the
most significant remaining positions. With proper bookkeeping, this algorithm requires
\BigO{N} time. Assuming that all moves are equally likely to be fitness improving
during enumeration, we argue below that the optimal bit ordering for skipping solutions is a greedy
ordering.

\subsection{All Optimal Orderings are Greedy}
The quality of an ordering is determined by how many solutions
Algorithm~\ref{alg-enumerate} can skip when using that ordering.
Each time the loop on Line~\ref{alg-enumerate-bincheck} stops with $index>0$,
$2^{index}$ solutions have been skipped. For practical purposes we will
assume that all moves are equally likely to be fitness improving.

@@

\begin{proof}
When Algorithm~\ref{alg-enumerate} is applied to a linear
function, each index of $move\_bin$ can contain at most 1
move, and a move is fitness improving whenever it disagrees with the global optimum.
Initially $index$ is set to $N-1$ and is decreased by at least 1 in any step after the first.
Each step will flip a bit which disagreed with the global optimum.
Before finding the global optimum no carry operations can occur, meaning the
cost of each iteration is equal to the amount by which $index$ is decreased.
Therefore, Algorithm~\ref{alg-enumerate} requires \BigO{N} time to reach
the global optimum.

@@

\section{Experiments}
We compare three methods for finding all local optima: Gray-Box, Hyper, and Hyper-Reorder.
The Gray-Box method refers to previous work~\cite{ochoa:2015:crossovernetworks} which does
not perform hyperplane elimination, with Hyper and Hyper-Reorder being hyperplane elimination
without and with the enumeration order improvements discussed in Section~\ref{sec-reorder}.

@@

\subsection{Finding Local Optima}
\label{sec-1-bit-optima}
Our first experiment is to determine the empirical complexity of each method with respect to problem size.
We tested 932 task configurations (problem, $N$, $r$), with 681 completed by at least one method.
Of those, Hyper-Reorder had the lowest
mean time to find all local optima in all but 2. In both of those cases Hyper-Reorder was within a tenth of a second
of the best method.
Figure~\ref{fig-boxplot-method} shows the variance in runtime for the largest problem size where all three methods
were successful. In all cases it is clear that Hyper-Reorder outperforms the other methods.

@@

@@

Perhaps most striking in Figure~\ref{fig-slope-method} is that on Adjacent NKq, Random NKq,
and Ising Spin Glass, increasing $r$ \emph{reduces} Hyper-Reorder's growth complexity.
Increasing $r$ creates more moves which can be fitness improving. As a result there
are more hyperplanes which can potentially be detected and eliminated.
On Concatenated Traps, no flip using $1 < r < k$ bits can be fitness improving
without a 1-bit flip being fitness improving, meaning no additional hyperplanes can be skipped.
For MAX-kSAT, increasing $r$ likely compounds the high dependency problems discussed in
Section~\ref{sec-1-bit-optima}.

@@

MAX-kSAT and Ising Spin Glasses generally appear to contain far more local optima than
any of the other problems we enumerated. With just $N=49$, Ising Spin Glasses average
nearly 800 million local optima. For comparison, Adjacent NKq with $N=49$ averages fewer than 300 thousand,
and with $N=77$ Adjacent NKq still only has fewer than 500 million.
As with NKq, the polynomially solvable Ising Spin
Glass problem had more 1-bit local optima than the NP-Hard MAX-kSAT. Unlike with NKq, this
relationship generally does not change as $r$ increases. However, both do still
see large decreases in the number and growth rate of local optima. With $N=49$
the number of 4-bit local optima Ising Spin Glass is on average under 70 thousand,
a reduction of four orders of magnitude.

>>>>

\begin{abstract}
Examining the properties of local optima is a common method for understanding
combinatorial-problem landscapes.
Unfortunately,
exhaustive methods for finding local optima are limited to very small problem sizes.
We propose a method for exploiting problem structure
to skip hyperplanes
that cannot contain local optima, allowing runtime to scale with
the number of local optima instead of with the landscape size.
We prove optimality for linear and $k$-bounded separable problems, and we provide
empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
We further refine this method to find only solutions
that cannot be improved by flipping $r$ or fewer bits, which counterintuitively
can reduce total runtime. While previous methods were limited to
landscapes with at most $2^{34}$ binary strings, hyperplane elimination can enumerate the same problems with
$2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
We believe this increase in size will help further our understanding of complex real-world landscapes.

@@

We use two common variants of NKq in our experiments which specify how the $K$ additional
variables in each subfunction are chosen: Adjacent and Random. In Adjacent NKq, $f_i$ depends
on variable indices $[i..(i+k) \bmod N]$. In Random NKq, $f_i$ depends on $x_i$ and a random
set of $K$ unique variables that does not include $x_i$. While Random NKq is NP-Hard,
the structure of Adjacent NKq allows for a polynomial time dynamic programming solution~\cite{wright:2000:solvingnk}.
Adjacent NKq is therefore often used to approximate a search algorithm's ability to
solve Random NKq.

@@

The first major result in Gray-Box optimization was the proof that the list
of fitness-improving moves from a solution can be updated after a bit flip in \BigO{1}
time~\cite{whitley:2012:constant}.
The fitness effect of flipping $x_j$ only changes after flipping $x_i$ if
there is a non-linear relationship between $x_i$ and $x_j$.
In an Mk Landscape, variables
can only have a non-linear relationship if they appear together in at least one $s_i$.
Therefore by definition the total number of non-linear
relationships in an Mk Landscape is linear with $N$, meaning on average the amortized
number of non-linear relationships per variable is \BigO{1}.

This result was extended, with no increase in asymptotic costs,
to include search for
solutions that cannot be improved by flipping $r$ or fewer bits~\cite{chicano:2014:ball}.
Consider two variables $x_i$ and $x_j$ that do not appear in the same $s_i$.
By definition the fitness effect of flipping both $x_i$ and $x_j$ is equal to the sum
of flipping each independently. As a result, if neither individual flip is fitness improving,
flipping both together cannot be fitness improving. By examining the graph of non-linear
interactions between variables, this principle can be extended to any collection of $r$
variables, with the number of useful collections growing at \BigO{N} when $r$ is a small constant.
As a result, the time to update the list of improving moves after up to $r$ bit-flips is still \BigO{1}.

@@

\section{Hyperplane Elimination}
\begin{figure}
  \centering
  \includegraphics[width=.75\columnwidth,height=\textheight,keepaspectratio]{Enumerate}
  %\includegraphicsfit{Enumerate}
  \caption{Example change of ordering. The gray variables represent all dependencies
           for move $m_i$ which flips variable $F$. By reordering, $m_i$'s lowest $index$ dependency improves from 2 to 4.}
  \label{fig-enumerate}
\end{figure}

Due to the limited non-linearity of the Gray-Box domain, it is possible to exclude large
parts of the search space without missing any local optima.
Consider the representation presented in the top of Figure~\ref{fig-enumerate}. In a Black-Box
domain, enumeration would progress as a binary counter, treating $index$ 0 (symbol $A$ in the solution) as
the least significant bit. This ordering ensures that before changing $index$ $i$, all possible settings of $index$
0 through $i-1$ have been tested. This corresponds to examining the hyperplane where the lowest $i-1$ positions
can vary and all other positions remain fixed. The Gray-Box domain makes it possible to skip hyperplanes
that cannot contain local optima. In Figure~\ref{fig-enumerate}, move $m_i$ which flips variable $F$ is a fitness improvement
when enumeration starts (all variables set to 0). Due to the known relationships between variables,
we know that the quality of $m_i$ only depends on variables $C$, $E$, $F$, and $H$.
Therefore, until one of those four variables are modified, the solution cannot be a local optimum.
As a result, the hyperplane **0*00*0 cannot contain a local optima and can
therefore be eliminated from consideration during enumeration.
More generally, if at any point during enumeration
there exists a fitness-improving move, no local optima can exist until at least one
dependency of that move is modified. Any hyperplane that has all of a fitness-improving
move's dependencies fixed cannot contain any local optima.

\begin{algorithm}
  \caption{Find all local optima using Hyperplane Elimination.}
  \label{alg-enumerate}
  \begin{algorithmic}[1]
    \Require $solution \leftarrow \{0\}^N$
    \Require $move\_bin$ (How many improving moves have each minimum dependency)
    \Require \textproc{MakeMove} (Function that flips $index$ in $solution$ and updates $move\_bin$)
    \Ensure $found$ (list of all local optima)
    \State $found \leftarrow [~]$
    \State $index \leftarrow N-1$
    \While{$index < N$}
      \While{$index > 0$ \textbf{and} $move\_bin[index] = 0$}\label{alg-enumerate-bincheck}
        \State $index \leftarrow index-1$
      \EndWhile
      \If{$index = -1$}
        \State $found \leftarrow found + [solution]$
        \State $index \leftarrow 0$
      \EndIf
      \While{$index < N$ \textbf{and} $solution[index] = 1$}\label{alg-enumerate-counter}
        \State \Call{MakeMove}{$index$}
        \State $index \leftarrow index + 1$
      \EndWhile
      \If{$index < N$}
        \State \Call{MakeMove}{$index$}\label{alg-enumerate-insert}
      \EndIf
    \EndWhile
  \end{algorithmic}
\end{algorithm}

@@

\section{Reordering Variables}
\label{sec-reorder}
As a final efficiency, the order in which variables in $solution$ are $index$ed can be reordered.
When a move is fitness-improving,
the amount of search space that is skipped depends on how high its lowest $index$ dependency is. Therefore,
by rearranging the order to make its lowest $index$ dependency higher, more search space can be skipped. 
Figure~\ref{fig-enumerate} shows how changing the $index$ order of variables
improves $m_i$'s lowest $index$ dependency from 2 to 4. Consider that before reordering $m_i$ allows the hyperplane
**000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering the hyperplane that can be skipped is ****0000.
Now, whenever $m_i$ is a fitness improvement, 4 times as many solutions are skipped.

We perform this reordering in a greedy fashion, such that the move with the
least-unmapped dependencies has all of its remaining dependencies mapped to the
most significant remaining positions. With proper bookkeeping, this algorithm requires
\BigO{N} time. We argue below that the optimal bit ordering for skipping solutions
is a greedy ordering.

The quality of an ordering is determined by how many solutions
Algorithm~\ref{alg-enumerate} can skip when using that ordering.
Each time the loop on Line~\ref{alg-enumerate-bincheck} stops with $index>0$,
$2^{index}$ solutions have been skipped. For practical purposes we will
assume that all moves are equally likely to be fitness improving.

@@

\begin{proof}
When Algorithm~\ref{alg-enumerate} is applied to a linear
function, an index $i$ of $move\_bin$ is non-zero if and only if
$x_i$ disagrees with the global optimum.
Initially $index$ is set to $N-1$ and is decreased by at least 1 in any step after the first.
Each step will flip a bit which disagreed with the global optimum.
Before finding the global optimum no carry operations can occur, meaning the
cost of each iteration is equal to the amount by which $index$ is decreased.
Therefore, Algorithm~\ref{alg-enumerate} requires \BigO{N} time to reach
the global optimum.

@@

\section{Experiments}
We compare three methods for finding all local optima: Gray-Box, Hyper, and Hyper-Reorder.
The Gray-Box method refers to previous work~\cite{ochoa:2015:crossovernetworks} which does
not perform hyperplane elimination, with Hyper and Hyper-Reorder being hyperplane elimination
without and with the order improvements discussed in Section~\ref{sec-reorder}.

@@

\subsection{Finding Local Optima}
\label{sec-1-bit-optima}

@@

%Our first experiment is to determine the empirical complexity of each method with respect to problem size.
We tested 932 task configurations (problem, $N$, $r$), with 681 completed by at least one method.
Of those, Hyper-Reorder had the lowest
mean time to find all local optima in all but 2. In both of those cases Hyper-Reorder was within a tenth of a second
of the best method.
Figure~\ref{fig-boxplot-method} shows the variance in runtime for the largest problem size where all three methods
were successful. In all cases it is clear that Hyper-Reorder outperforms the other methods.

@@

Perhaps most striking in Figure~\ref{fig-slope-method} is that on Adjacent NKq, Random NKq,
and Ising Spin Glass, increasing $r$ \textbf{reduces} Hyper-Reorder's growth complexity.
Increasing $r$ creates more moves which can be fitness improving. As a result there
are more hyperplanes which can potentially be detected and eliminated.
On Concatenated Traps, no flip using $1 < r < k$ bits can be fitness improving
without a 1-bit flip being fitness improving, meaning no additional hyperplanes can be skipped.
For MAX-kSAT, increasing $r$ likely compounds the high dependency problems discussed in
Section~\ref{sec-1-bit-optima}.

@@

MAX-kSAT and Ising Spin Glasses generally appear to contain far more local optima than
any of the other problems we enumerated. With just $N=49$, Ising Spin Glasses average
nearly 800 million local optima. For comparison, Adjacent NKq with $N=49$ averages fewer than 300 thousand,
and with $N=77$ Adjacent NKq still only has fewer than 500 million.
As with NKq, the polynomially solvable Ising Spin
Glass problem had more 1-bit local optima than the NP-Hard MAX-kSAT. Unlike with NKq, this
relationship generally does not change as $r$ increases. However, both do still
see large decreases in the number and growth rate of local optima. With $N=49$
the number of 4-bit local optima in an Ising Spin Glass is on average under 70 thousand,
a reduction of four orders of magnitude.
